---
title: "Iron Depletion effect on cell lines"
output: html_notebook
---

```{r include = FALSE}
library(GEOquery)
library(magrittr)
library(dplyr)
library(ggplot2)
library(here)
library(AnnotationHub)
```

## 0. Introduction

The purpose of this analysis is to examine gene expression data from a pure 
cell line that has undergone iron depletion, and see what's happening 
in terms of IRE gene enrichment. 

We will be using a dataset from GEO with accession number 
[GSE32369](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE32369). 
The title of this study is *Effect of iron chelators on global gene 
expression pattern in SW480 cells*, where the SW480 cell line is derived from 
colon cancer. Unfortunately I couldn't find any 
RNA-seq datasets so this one is also a microarray dataset 

The summary of the dataset is:

<blockquote>
Gene expression analysis of SW480 cells treated with inhibitor compounds for 
6 hours. Results provide insights into the role of iron in Wnt signalling 
and demonstrate that iron depletion is the primary mode of actions of these 
compounds on Wnt pathway.

SW480 cells were incubated with 10 mM of compounds (OICR623), 
control (DMSO) and two known iron chelators (DFO and Deferasirox) for 
6 hours. RNA was extracted and cDNA samples isolated from 2-4 independent 
experiments were hybridized to the Affymetrix GeneChip® Human Gene 1.0 ST array.
</blockquote>

The iron chelators DFO and Deferasirox are relevant to our work. 


## 1. Data Import

The expression and sample data was imported from GEO as follows:

```{r eval=FALSE}
counts <- getGEO("GSE32369", GSEMatrix = TRUE)

samples <- counts$GSE32369_series_matrix.txt.gz %>% 
  phenoData() %>% 
  pData %>% 
  as_tibble

counts <- counts$GSE32369_series_matrix.txt.gz %>%
  exprs() 
```

Preview the data:

```{r}
counts %>% as.data.frame %>% rownames_to_column("gene_id") %>% as_tibble
samples
```

We will clean up the sample metadata and keep the relevant columns only. 

```{r}
samples %<>% dplyr::select(title, 
                          geo_accession,
                          characteristics_ch1.2) %>% 
  dplyr::rename(
    sample = title, 
    id = geo_accession, 
    treatment = characteristics_ch1.2
    ) %>%
  dplyr::mutate(
    treatment = gsub(x=treatment,
                     pattern = "^.* |\\(|\\)", 
                     replacement=""),
    treatment = as.factor(treatment)
  ) %>%
  dplyr::arrange(treatment) 
```
```{r}
samples
```

Check the "id" column in `samples` corresponds to the `sampleNames`. 
```{r}
sampleNames <- colnames(counts)
samples$id %in% sampleNames # Are all samples in sample table present in counts samples?
samples$id == sampleNames # Is the order the same?
```

The ordering of samples isn't the same across the sample table and the 
counts matrix, so we'll reorder the counts. 
```{r}
counts <- counts[, samples$id]
```
```{r}
sampleNames <- colnames(counts)
samples$id %in% sampleNames # Are all samples in sample table present in counts samples?
samples$id == sampleNames # Is the order the same?
```



## 2. Analysis

We are really only interested in whether IRE gene enrichment is present. 
However, to do this, we need to set up a `voom` obect *ala* **limma**. 

### 2.1. Gene annotation

```{r}
ah <- AnnotationHub()
ah %>%
  subset(grepl("sapiens", species)) %>%
  subset(rdataclass == "EnsDb")
```

```{r}
ensDb <- ah[["AH64923"]]
genes <- genes(ensDb)
genes <- as.data.frame(genes)
head(genes)


```

```{r}
rownames(counts) %>% head
```

Before annotation, we have to map the probe IDs to their annotations. 

```{r}
probeAnnot <- read_csv("~/Downloads/HuGene-1_0-st-v1.na36.hg19.probeset.csv/HuGene-1_0-st-v1.na36.hg19.probeset.csv", comment = "#")
```
```{r}
head(probeAnnot)

rownames(counts) %in% probeAnnot$transcript_cluster_id %>% sum
```

```{r}
genes <- data.frame(transcript_cluster_id = as.numeric(rownames(counts))) %>%
  left_join(probeAnnot, by = "transcript_cluster_id")

genes %>% tail(10000) %>% View
```

## 3. Process raw probe intensities

Steps

- Read in raw probe cel file into an `Affybatch` using `read.affybatch()`.
- Use `expresso` to get from raw probe intensities to expression values. 

```{r}

affybatch <- ReadAffy(celfile.path = here::here("R/IronDepleted/data/GSE32369_RAW"))

normData <- expresso(affybatch, 
                     bgcorrect.method = "rma", 
                     normalize.method = "constant", 
                     pmcorrect.method="pmonly", 
                     summary.method="avgdiff")
```

The expression matrix looks like this:
```{r}
normData@assayData$exprs %>% head
```

Add some sample metadata:
```{r}
normData@phenoData@data %<>% 
  rownames_to_column("filename") %>%
  mutate(id = gsub(x = filename, pattern = ".CEL.gz", replacement = "")) %>%
  left_join(samples, by="id") %>%
  dplyr::arrange(treatment) %>%
  column_to_rownames("filename") 
```

```{r}
#library(hugene10stv1cdf)
library(hugene10sttranscriptcluster.db)
rownames(counts) %in% keys(hugene10sttranscriptcluster.db) %>% summary


annot <- AnnotationDbi::select(
  x = hugene10sttranscriptcluster.db, 
  keys = rownames(counts),
  columns = c("PROBEID", "ENSEMBL", "ENTREZID", "SYMBOL"),
  keytype = "PROBEID"
)

all(rownames(counts) %in% annot$PROBEID) #TRUE
 
dup.ids <- annot$PROBEID[duplicated(annot$PROBEID)] %>% 
  unique %>%
  sort
length(dup.ids) #2916

# Example probe
annot[annot$PROBEID == dup.ids[2], ]

# To collapse down the duplicated genes, we will concatenate 
# the multiple genes for one probe 
collapser <- function(x){
  x %>% unique %>% sort %>% paste(collapse = "|")
}
annot <- annot %>% 
  group_by(PROBEID) %>%
  summarise_each(funs(collapser)) %>% 
  ungroup
# http://biolearnr.blogspot.com/2017/05/bfx-clinic-getting-up-to-date.html

# Filter out rows which don't have ensembl IDs. 
annot %<>% dplyr::filter(!ENSEMBL == "")

head(annot)

# We also would want to restrict the expression data to these probes:
exprs <- normData@assayData$exprs %>% as.data.frame %>%
  rownames_to_column("PROBEID") %>% 
  dplyr::filter(PROBEID %in% annot$PROBEID) %>%
  column_to_rownames("PROBEID") %>%
  magrittr::extract(annot$PROBEID, )

# Check all true
(rownames(exprs)==annot$PROBEID) %>% summary #TRUE

# add annotation data to the feat
```



### 2.2. DGEList

The `samples`, `counts` and `genes` will be stored in a DGEList object 
for the DE analysis using *limma*. 

```{r}
dge <- DGEList(
  counts = counts,
  samples = samples,
  genes = genes,
  remove.zeros = TRUE
) %>%
  calcNormFactors("TMM")
```