---
title: "R Notebook"
output: html_notebook
---

```{r Setup, include = FALSE}
library(readr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(hgug4110b.db)
library(UpSetR)
library(GEOquery)
library(tibble)
library(here)
library(ggrepel)
library(reshape2)
library(GSEABase)
library(dplyr)
library(fgsea)
library(org.Hs.eg.db)
library(org.Dr.eg.db)
library(biomaRt)
library(limma)
library(stringr)
library(openxlsx)
library(pheatmap)
theme_set(theme_bw())

write_fit <- function(fit, results=NULL, F.adjust="none", adjust="fdr", method="global", ...) {
    if (!is(fit, "MArrayLM")) 
        stop("fit should be an MArrayLM object")
    if (!is.null(results) && !is(results, "TestResults")) 
        stop("results should be a TestResults object")
    if (is.null(fit$t) || is.null(fit$p.value)) 
        fit <- eBayes(fit)
    method <- match.arg(method, c("separate", "global"))
    p.value <- as.matrix(fit$p.value)
    if (adjust == "none") {
        p.value.adj <- NULL
    }
    else {
        p.value.adj <- p.value
        if (method == "separate") 
            for (j in 1:ncol(p.value)) p.value.adj[, j] <- p.adjust(p.value[, 
                j], method = adjust)
        if (method == "global") 
            p.value.adj[] <- p.adjust(p.value, method = adjust)
    }
    if (F.adjust == "none" || is.null(fit$F.p.value)) 
        F.p.value.adj <- NULL
    else F.p.value.adj <- p.adjust(fit$F.p.value, method = F.adjust)
    
    tab <- list()
    tab$A <- (fit$Amean)
    tab$Coef <- (fit$coef)
    tab$t <- (fit$t)
    tab$p.value <- (p.value)
    tab$p.value.adj <- (p.value.adj)
    tab$F <- (fit$F)
    tab$F.p.value <- (fit$F.p.value)
    tab$F.p.value.adj <- F.p.value.adj
    tab$Res <- unclass(results)
    tab$Genes <- fit$genes
    tab <- data.frame(tab, check.names = FALSE)
    return(tab)
}
```

# Iron Overload

## 1. Import Dataset

The processed data was downloaded from GEO 
([GSE3573](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE3573)). 
This has the expression already summarised to probe-set level. 
I wanted the raw probe intensities, but when I tried to download this from 
GEO, there was only an annotation text file in the .tar. No raw data files 
unfortunately. 

**Summary of the study**:
<blockquote>
We investigated transcriptional response of CaCo-2 cells to iron treatments, we studied hemin effect by adding hemin to DMEM-FBS medium and iron deficiency effects in using an iron free medium compared to the same supplemented with FAC (ferric ammonium citrate).
Keywords: various iron treatment, differential gene expression, hemin treatment, iron-free
</blockquote>

```{r}
geo <- getGEO("GSE3573", GSEMatrix = TRUE)
```

```{r}
exprs <- geo$GSE3573_series_matrix.txt.gz@assayData$exprs
samples <- geo$GSE3573_series_matrix.txt.gz@phenoData@data
annot <- geo$GSE3573_series_matrix.txt.gz@featureData@data
dim(annot)
dim(exprs)
```

The processed expression matrix and sample and probe annotations were extracted 
from the GEO object. There are `r nrow(annot)` features (probesets) in this 
dataset.

The expression has already been log2 transformed:

```{r}
exprs %>% 
  as.data.frame %>%
  melt %>% 
  group_by(variable) %>%
  dplyr::summarise(min = min(value), 
                   max = max(value))
```


A bit of formatting: 
```{r}
exprs %<>% set_rownames(annot$NAME)
samples %<>% dplyr::select(title, 
                           geo_accession, 
                           treatment_protocol_ch1) %>%
  dplyr::mutate(treatment = c(rep("DMEM_FBS_Medium", 3),
                              rep("DMEM_FBS_Medium_and_Hemin", 3), 
                              rep("Ironfree_Medium", 3),
                              rep("Ironfree_Medium_and_FAC", 3))) 
  #dplyr::select(title, geo_accession, treatment)

samples
```

## 2. Principal Component Analysis
```{r}
pca <- exprs %>% t %>% prcomp()
summary(pca)
pca_plot_GSE3573 <- pca$x %>% as.data.frame %>% 
  rownames_to_column("geo_accession") %>% 
  left_join(samples, by = "geo_accession") %>%
  ggplot(aes(x = PC1, y = PC2, colour = treatment)) + 
  geom_point(size=3) + 
  ggrepel::geom_text_repel(aes(label = geo_accession)) +
  theme(aspect.ratio = 1) +
  labs(x = "Principal Component 1 (24.8%)", 
       y = "Principal Component 2 (13.8%)",
       colour = "Treatment")
pca_plot_GSE3573
#export::graph2ppt(pca_plot_GSE3573, here::here("R/IronDepleted/fig/pca_plot_GSE3573"))
#saveRDS(pca_plot_GSE3573, here::here("R/IronDepleted/fig/pca_plot_GSE3573.rds"))
#pca_plot_GSE3573 <- readRDS(here::here("R/IronDepleted/fig/pca_plot_GSE3573.rds"))
```

## 3. Annotation
```{r include = FALSE, eval=FALSE}
# Check that the probe IDs are in the annotation package 
rownames(exprs) %in% keys(hgug4110b.db) %>% summary  # all TRUE

# Retrieve gene id annotations
annot <- AnnotationDbi::select(
  x = hgug4110b.db, 
  keys = rownames(exprs),
  columns = c("PROBEID", "ENSEMBL", "ENTREZID", "SYMBOL"),
  keytype = "PROBEID"
)
all(rownames(exprs) %in% annot$PROBEID)  # TRUE

# Check for duplicate probe ids (assigned to different genes)
dup.ids <- annot$PROBEID[duplicated(annot$PROBEID)] %>% 
  unique %>%
  sort
length(dup.ids)  # 4430

# Example probe
annot[annot$PROBEID == dup.ids[2], ]

# Collapse down the duplicated genes so each row is 
# a unique probe ID
collapser <- function(x){
  x %>% unique %>% sort %>% paste(collapse = "|")
}
annot <- annot %>% 
  group_by(PROBEID) %>%
  summarise_each(funs(collapser)) %>% 
  ungroup

# Filter out probes which don't have ensembl IDs
annot %<>% dplyr::filter(!ENSEMBL == "")

head(annot)
dim(annot) # 43,285 probes

# We also would want to restrict the expression data to these probes
# and get the ordering to match
exprs <- normData@assayData$exprs %>% as.data.frame %>%
  rownames_to_column("PROBEID") %>% 
  dplyr::filter(PROBEID %in% annot$PROBEID) %>%
  column_to_rownames("PROBEID") %>%
  magrittr::extract(annot$PROBEID, )

# Check the ordering of rownames is exactly the same
(rownames(exprs)==annot$PROBEID) %>% summary  # all TRUE
```



## 4. DE analysis

```{r}
# LIMMA ANALYSIS -------------------------------------------------------------------------------------------------------
# Bg: Need the following objects from limma analysis to perform IRE enrichment analysis:
#     voomData, design, contrasts

design <- model.matrix(~0 + treatment, data = samples) %>%
  set_colnames(gsub(x = colnames(.), pattern = "treatment", replacement = ""))

contrasts <- makeContrasts(
  levels = colnames(design), 
  IronOverload = DMEM_FBS_Medium_and_Hemin - DMEM_FBS_Medium,
  IronDeficiency = Ironfree_Medium - Ironfree_Medium_and_FAC
)

# Estimate sample-level weights using the groups defined in design matrix.
# weights_array <- arrayWeights(exprs, design=design, method="reml") %>% 
#   set_names(colnames(exprs))

fit <- lmFit(exprs, design) %>%
  contrasts.fit(contrasts)%>%
  eBayes()

hist(fit$Amean) # Many genes have low expression

results <- decideTests(fit, 
                       p.value = 0.05, 
                       adjust.method = "fdr", 
                       method="global")

results %>% summary()

fitresults <- write_fit(fit, results) %>%
  as.data.frame %>%
  rownames_to_column("NAME") %>%
  dplyr::left_join(annot %>% 
                     dplyr::select(ID, 
                                   NAME, 
                                   GENE,
                                   GENE_SYMBOL,
                                   GENE_NAME, 
                                   ENSEMBL_ID), by = "NAME")

vennCounts(results)

volcanoplot(fit, coef= "IronOverload")

```



### 4.1. Volcano Plot

- The following volcano plots indicate the genes determined to be DE 
(FDR-adjusted p-value < 0.05) under iron overload and iron deficiency 
treatments. 

```{r volcano-plots}
# iron overload
volcanoPlot_ironOverload <- fitresults %>% 
  dplyr::select(NAME, Coef.IronOverload, 
                p.value.IronOverload,
                p.value.adj.IronOverload,
                Res.IronOverload) %>% 
  as_tibble %>%
  set_colnames(c("NAME", "logFC", "p", "fdr", "Significant")) %>%
  dplyr::distinct(.keep_all = TRUE)  %>%
  dplyr::arrange(p) %>%
  ggplot(aes(x = logFC, y = -log10(p))) +
  geom_point(aes(colour = as.factor(Significant)), alpha=0.5) +
  scale_colour_manual(values = c("red", "#cccccc", "blue")) +
  #ggrepel::geom_text_repel(aes(label = ifelse(Significant != 0, NAME,""))) +
  ggtitle("Volcano plot of Iron Overload") +
  labs(x = "log2 fold change", y = "-log10(p value)") +
  geom_hline(yintercept=-log10(3.515036e-04), linetype="solid", 
                color = "#555555", size=1) +
  theme(legend.position = "none",
        aspect.ratio = 1)
volcanoPlot_ironOverload %>% graph2pdf(here::here("R/IronDepleted/fig/volcanoPlot_ironOverload"))


# iron deficiency
volcanoPlot_ironDeficiency <- fitresults %>% 
  dplyr::select(NAME, Coef.IronDeficiency, 
                p.value.IronDeficiency,
                p.value.adj.IronDeficiency,
                Res.IronDeficiency) %>% 
  as_tibble %>%
  set_colnames(c("NAME", "logFC", "p", "fdr", "Significant")) %>%
  dplyr::distinct(.keep_all = TRUE)  %>%
  dplyr::arrange(p) %>% 
  ggplot(aes(x = logFC, y = -log10(p))) +
  geom_point(aes(colour = as.factor(Significant)), alpha=0.5) +
  scale_colour_manual(values = c("red", "#cccccc", "blue")) +
  #ggrepel::geom_text_repel(aes(label = ifelse(Significant != 0, NAME,""))) +
  ggtitle("Volcano plot of Iron Deficiency") +
  labs(x = "log2 fold change", y = "-log10(p value)") +
  geom_hline(yintercept=-log10(3.544330e-04), linetype="solid", 
                color = "#555555", size=1) +
  theme(legend.position = "none",
        aspect.ratio = 1)
volcanoPlot_ironDeficiency %>% export::graph2pdf(here::here("R/IronDepleted/fig/volcanoPlot_ironDeficiency"))

```
```{r}
volcanoPlot_ironDeficiency
volcanoPlot_ironOverload
```

## 5. Gene set enrichment tests

- We will perform enrichment testing on the following gene sets:

    - Hallmark collection from MSigDB
    - Human predicted IRE gene sets
    
- First we need to format the expression matrix properly. 
At the moment, the rownames of the expression matrix don't correspond to 
only one type of gene ID (some RefSeq, some probes, etc.). Although we are 
going to have to lose some information, we will collapse them down to 
representative Ensembl IDs. 

- Instead of using the `voom` object from *limma*, because this is a 
microarray dataset we will need to use the `fit` object and expression object. 

```{r}
# Combine the expression data with annotations
gene2trans <- as.data.frame(org.Hs.egENSEMBLTRANS) %>% 
  left_join(as.data.frame(org.Hs.egENSEMBL))%>%
  set_colnames(c("ENTREZ", "ENSEMBL_ID", "ENSEMBL"))
annot %<>% left_join(gene2trans, by = "ENSEMBL_ID")

datET <- exprs %>% 
  as.data.frame %>% 
  rownames_to_column("NAME") %>%
  left_join(annot, by = "NAME") %>% 
  mutate(GENE = as.character(GENE)) %>%
  left_join(gene2trans, by = c("GENE"="ENTREZ")) %>%
  dplyr::filter(ENSEMBL != "" | !is.na(ENSEMBL)) %>%
  dplyr::distinct(ENSEMBL, .keep_all = TRUE) %>%
  set_rownames(.$ENSEMBL)

# Collapse rows
collapsed <- WGCNA::collapseRows(datET[,2:13], 
                          rowGroup = datET$ENSEMBL, 
                          rowID = rownames(datET), 
                          method ="MaxMean")

colExprs <- collapsed$datETcollapsed
colExprs <- datET[,2:13]
 dim(colExprs)  #  7,793 genes remaining

colAnnot <- data.frame(ENSEMBL = rownames(colExprs)) %>%
  left_join(annot, by = "ENSEMBL") %>%
  dplyr::distinct(ENSEMBL, .keep_all = TRUE)

newFit <- lmFit(colExprs, design) %>%
  contrasts.fit(contrasts) %>%
  eBayes()

results <- decideTests(newFit, 
                       p.value = 0.05, 
                       adjust.method = "fdr",
                       method = "global")

results %>% summary()
```

- The number of DE genes has been lowered due to cutting out a lot of genes, but 
we'll proceed ahead anyway to see what the results look like. 

- We will perform enrichment testing in a combined way by putting all of the 
gene sets together. 

```{r}

source(here::here("R/GSEA/combinedGSEA_ma.R"))
humanIreGenes <- readRDS(here::here("R/IREGenes/data/human_ireGenes.rds"))

genesetsDir <- here::here("R", "GSEA", "data", "human")
h_mapped <- readRDS(file.path(genesetsDir, "ens_h_mapped.rds"))


# Build index specific to the ordering of the rownames in the expression matrix
humanIreIdx <- limma::ids2indices(humanIreGenes, 
                                  identifiers = rownames(colExprs))

humanHallmarkIdx <- limma::ids2indices(h_mapped, identifiers = rownames(colExprs))

combinedIdx <- c(humanIreIdx, humanHallmarkIdx)
  
ireEnr <- combinedGSEA_ma(colExprs, 
                          fit = newFit, 
                          design = design, 
                          contrasts = contrasts, 
                          idx = humanIreIdx)

hEnr <- combinedGSEA_ma(colExprs, 
                          fit = newFit, 
                          design = design, 
                          contrasts = contrasts, 
                          idx = combinedIdx)
# Results
ireEnr
hEnr

# Export
ireEnr %>% saveRDS(here::here("R/IronDepleted/data/GSE3573_ireEnr.rds"))
hEnr %>% saveRDS(here::here("R/IronDepleted/data/GSE3573_hEnr.rds"))
```

### 5.1. Butterfly plot

- The GSEA results can be visualised using a back-to-back barchart. 

```{r fig.height=17, fig.width=20}
# extract up/downregulated gene proportions from mroast
x_props <- hEnr$indivTest$IronOverload$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="IronOverload",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

# join to fry results
x_fry_with_props <- hEnr$indivTest$IronOverload$fry %>%
  #dplyr::filter(FDR.Mixed < 0.05) %>%
  dplyr::select(Geneset) %>%
  left_join(x_props)

x_fry_with_props

# extract proportions from mroast
y_props <- hEnr$indivTest$IronDeficiency$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="IronDeficiency",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

# join to fry results
y_fry_with_props <- hEnr$indivTest$IronDeficiency$fry %>%
  #dplyr::filter(FDR.Mixed < 0.05) %>%
  dplyr::select(Geneset) %>%
  left_join(y_props)

y_fry_with_props

# create the plot
# adapted from https://stackoverflow.com/questions/4559229/drawing-pyramid-plot-using-r-and-ggplot2
butterfly <- bind_rows(x_fry_with_props, y_fry_with_props) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, pattern = "_", replacement = " "),
                Geneset = gsub(x = Geneset, pattern = "HALLMARK", replacement = "")) %>%
  reshape2::melt(id.vars=c("Geneset","NGenes","contrast")) %>%
  dplyr::arrange(desc(NGenes)) %>% 
  dplyr::rename(direction = variable, ngenes = value) %>%
  dplyr::mutate(Geneset = factor(Geneset, levels = unique(Geneset)),
                direction = factor(direction, levels = c("notSig", "down", "up")),
                contrast = factor(contrast, levels = c("IronOverload","IronDeficiency"))) %>%
  ggplot(aes(x = Geneset, y = ngenes, fill = direction)) +
  geom_col(width=0.6) +
  ggpol::facet_share(~contrast, scales = "free") +
  scale_fill_manual(values=c("#DDDDDD","cornflowerblue","#FE4365"))+
  coord_flip(ylim=c(0,300)) +
  theme(axis.line.y = element_blank())


butterfly

export::graph2ppt(butterfly, here("R/IronDepleted/fig/butterfly"))
```


## Barplot

```{r}
iron_barplot <- bind_rows(IronOverload = ireEnr$indivTest$IronOverload$mroast, 
          IronDef = ireEnr$indivTest$IronDeficiency$mroast, .id = "id") %>% 
  dplyr::select(id, Geneset, NGenes, PropDown, PropUp) %>% 
  dplyr::mutate(up = NGenes*PropUp, 
                down = NGenes*PropDown, 
                notSig = NGenes-up-down) %>% 
  dplyr::select(-NGenes, -PropDown, -PropUp) %>%  
  melt %>% 
  ggplot(aes(x = Geneset, y = value, fill = variable)) + 
  geom_col() + 
  ggpol::facet_share(~id) + 
  coord_flip() + 
  scale_fill_manual(values=c("#FE4365","cornflowerblue","#DDDDDD")) + 
  theme(aspect.ratio = 1)

saveRDS(iron_barplot, here::here("R/IronDepleted/fig/iron_barplot.rds"))
export::graph2ppt(iron_barplot, here::here("R/IronDepleted/fig/iron_barplot"))
```


## UpSet Plot

- We wish to prepare an UpSet plot showing the overlap in the leading-edge 
genes between IRE gene sets in both the **iron overload** and 
**iron deficiency** treatments. 

- First we need to prepare lists of leading-edge genes for each gene set.

```{r}
le_ire3_all <- list(
  IronOverload = hEnr$indivTest$IronOverload$fgsea %>%
  dplyr::filter(Geneset == "ire3_all") %>% 
  .$leadingEdge %>%
  .[[1]],
  IronDeficiency = hEnr$indivTest$IronDeficiency$fgsea %>%
  dplyr::filter(Geneset == "ire3_all") %>% 
  .$leadingEdge %>%
  .[[1]]
)

le_ire5_all <- list(
  IronOverload = hEnr$indivTest$IronOverload$fgsea %>%
  dplyr::filter(Geneset == "ire5_all") %>% 
  .$leadingEdge %>%
  .[[1]],
  IronDeficiency = hEnr$indivTest$IronDeficiency$fgsea %>%
  dplyr::filter(Geneset == "ire5_all") %>% 
  .$leadingEdge %>%
  .[[1]]
)

le_ire3_hq <- list(
  IronOverload = hEnr$indivTest$IronOverload$fgsea %>%
  dplyr::filter(Geneset == "ire3_hq") %>% 
  .$leadingEdge %>%
  .[[1]],
  IronDeficiency = hEnr$indivTest$IronDeficiency$fgsea %>%
  dplyr::filter(Geneset == "ire3_hq") %>% 
  .$leadingEdge %>%
  .[[1]]
)

le_ire5_hq <- list(
  IronOverload = hEnr$indivTest$IronOverload$fgsea %>%
  dplyr::filter(Geneset == "ire5_hq") %>% 
  .$leadingEdge %>%
  .[[1]],
  IronDeficiency = hEnr$indivTest$IronDeficiency$fgsea %>%
  dplyr::filter(Geneset == "ire5_hq") %>% 
  .$leadingEdge %>%
  .[[1]]
)

leadingEdges <- list(
  ire3_all = le_ire3_all,
  ire5_all = le_ire5_all,
  ire3_hq = le_ire3_hq,
  ire5_hq = le_ire5_hq
)

str(leadingEdges)
```


```{r old}
# leadingEdge_ire3all <- ireEnr$indivTest %>% lapply(function(x){
#   x$fgsea %>% 
#     dplyr::arrange(Geneset)
#  x$fgsea$leadingEdge[[1]]
# })
# 
# leadingEdge_ire5all <- ireEnr$indivTest %>% lapply(function(x){
#   x$fgsea %>% 
#     dplyr::arrange(Geneset)
#  x$fgsea$leadingEdge[[2]]
# })
# 
# leadingEdge_ire3hq <- ireEnr$indivTest %>% lapply(function(x){
#   x$fgsea %>% 
#     dplyr::arrange(Geneset)
#  x$fgsea$leadingEdge[[3]]
# })
# 
# leadingEdge_ire5hq <- ireEnr$indivTest %>% lapply(function(x){
#   x$fgsea %>% 
#     dplyr::arrange(Geneset)
#  x$fgsea$leadingEdge[[4]]
# })
# 
# leadingEdges <- list(
#   ire3_all = leadingEdge_ire3all,
#   ire5_all = leadingEdge_ire5all,
#   ire3_hq = leadingEdge_ire3hq,
#   ire5_hq = leadingEdge_ire5hq
# )
# 
# str(leadingEdges)
```

```{r}
l2m <- leadingEdges %>% lapply(fromList)
```

```{r}
upset_ire3all <- upset(l2m$ire3_all,
      nintersects = NA,
      empty.intersections = TRUE
      )
upset_ire3all 

graph2ppt(upset_ire3all, here("R/IronDepleted/fig/upset_ire3all"))

```

```{r}
upset_ire5all <- upset(l2m$ire5_all,
      nintersects = NA,
      empty.intersections = TRUE
      )
upset_ire5all

graph2ppt(upset_ire5all, here("R/IronDepleted/fig/upset_ire5all"))
```

