---
title: "Colour palettes"
output: html_notebook
---

```{r}
library(ggplot2)
library(leaflet)
library(RColorBrewer)
```

## Background

The purpose of this is to generate colour scales that can be used to 
represent the differences between *p*-values for some figures in the paper. 
The colour makes it easier to compare at a glance different results 
and see which ones were highly significant compared to those which only 
just passed the cutoff of 0.05. 

## Palette

First we need to calculate -log2(pvalue) for the IRE enrichment results. 
This spreads the numbers out more evenly, which will be better for scaling 
with the colour palette:

```{r}
gseaResults_mouse_ire <- readRDS(here::here("R/Mouse/data/gseaResults_mouse_ire.rds"))

df <- gseaResults_mouse_ire$combTest %>% bind_rows %>% 
  dplyr::select(id, Geneset, wilkinsonp, bonferroni) %>%
  dplyr::mutate(logP = -log2(wilkinsonp)) %>%
  mutate(sig = bonferroni < 0.05) %>%
  dplyr::filter(Geneset %in% c("ire3_all","ire5_all")) %>% as.data.frame

df$logP
```

The colour palette that will be used is the yellow-orange-red one. This is 
a discrete colour palette from **RColorBrewer**. 

```{r}
ylOrRd <- brewer.pal(n = 9, name = "YlOrRd")
display.brewer.pal(9, "YlOrRd")
```

The `colorNumeric` function interpolates a discrete palette to make it 
continuous. We also have to input the range of values which the palette 
can accept. This is arbitrary, but the range has to include all the 
values in `df$logP`. 

```{r}
range(df$logP)
```

We will use the range 0-20.

```{r}
pal <- colorNumeric(ylOrRd, 0:20)
```

The hex colour codes can be generated as follows
```{r}
pal(c(df$logP)) 
```

We can also preview these colours using (opens in Viewer tab).  
```{r}
previewColors(pal, df$logP)
```

