---
title: "Promoter Over-representation Analysis Results"
output: html_notebook
---

```{r Setup, include=FALSE}
library(here)
library(readr)
library(dplyr)
library(magrittr)
```


## Zebrafish

- Import in the known motif enrichment results from HOMER:

```{r}
zebrafish_files <- list.files(here::here("R/PromoterAnalysis/zebrafish"), 
                              recursive=TRUE, 
                              pattern="knownResults.txt", 
                              full.names=TRUE)

zebrafish_file_names <- list.files(here::here("R/PromoterAnalysis/zebrafish"), 
                              recursive=TRUE, 
                              pattern="knownResults.txt", 
                              full.names=FALSE) %>%
  gsub(x=., pattern = "/knownResults.txt", replacement = "")

zebrafish_files
zebrafish_file_names
```

- Bind rows for the four runs using HOMER with the `ire3_all`, `ire5_all`, 
`ire3_hq` and `ire5_hq` gene sets, and then perform p-value correction on 
all raw p-values. 

```{r}
zebrafish_res <- zebrafish_files%>%
  lapply(read_tsv) %>%
  set_names(zebrafish_file_names) %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::rename(pvalue = `P-value`) %>%
  dplyr::mutate(bon = p.adjust(pvalue, "bonferroni"),
                fdr = p.adjust(pvalue, "fdr")) %>%
  dplyr::arrange(fdr)
```

- As shown in the table below, no zebrafish IRE gene sets are enriched in 
known zebrafish transcription factor binding motifs (FDR < 0.05). 

```{r}
head(zebrafish_res, 20) %>% dplyr::select(Geneset, 
                                          `Motif Name`, 
                                          Consensus,
                                          fdr)
```


## Human

- Import in the known motif enrichment results from HOMER:

```{r}
human_files <- list.files(here::here("R/PromoterAnalysis/human"), 
                              recursive=TRUE, 
                              pattern="knownResults.txt", 
                              full.names=TRUE)

human_file_names <- list.files(here::here("R/PromoterAnalysis/human"), 
                              recursive=TRUE, 
                              pattern="knownResults.txt", 
                              full.names=FALSE) %>%
  gsub(x=., pattern = "/knownResults.txt", replacement = "")

human_files
human_file_names
```

- Bind rows for the four runs using HOMER with the `ire3_all`, `ire5_all`, 
`ire3_hq` and `ire5_hq` gene sets, and then perform p-value correction on 
all raw p-values. 

```{r}
human_res <- human_files%>%
  lapply(read_tsv) %>%
  set_names(human_file_names) %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::rename(pvalue = `P-value`) %>%
  dplyr::mutate(bon = p.adjust(pvalue, "bonferroni"),
                fdr = p.adjust(pvalue, "fdr")) %>%
  dplyr::arrange(fdr)
```

- No human IRE gene sets are significantly enriched for other TF 
binding motifs. 

```{r}
head(human_res, 20) %>% dplyr::select(Geneset, 
                                          `Motif Name`, 
                                          Consensus,
                                          fdr)
```



## Mouse

- Import in the known motif enrichment results from HOMER:

```{r}
mouse_files <- list.files(here::here("R/PromoterAnalysis/mouse"), 
                              recursive=TRUE, 
                              pattern="knownResults.txt", 
                              full.names=TRUE)

mouse_file_names <- list.files(here::here("R/PromoterAnalysis/mouse"), 
                              recursive=TRUE, 
                              pattern="knownResults.txt", 
                              full.names=FALSE) %>%
  gsub(x=., pattern = "/knownResults.txt", replacement = "")

mouse_files
mouse_file_names
```

- Bind rows for the four runs using HOMER with the `ire3_all`, `ire5_all`, 
`ire3_hq` and `ire5_hq` gene sets, and then perform p-value correction on 
all raw p-values. 

```{r}
mouse_res <- mouse_files%>%
  lapply(read_tsv) %>%
  set_names(mouse_file_names) %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::rename(pvalue = `P-value`) %>%
  dplyr::mutate(bon = p.adjust(pvalue, "bonferroni"),
                fdr = p.adjust(pvalue, "fdr")) %>%
  dplyr::arrange(fdr)
```

- No mouse IRE gene sets are significantly enriched for other TF 
binding motifs. 

```{r}
head(mouse_res, 20) %>% dplyr::select(Geneset, 
                                          `Motif Name`, 
                                          Consensus,
                                          fdr)
```

## Export Results

- The results are exported as an Excel notebook for inclusion in the 
Supplementary Information. 

```{r}
results <- list(
  human = human_res,
  mouse = mouse_res,
  zebrafish = zebrafish_res
)

openxlsx::write.xlsx(results, here("R/PromoterAnalysis/results/results.xlsx"))
```


## Session Info

```{r}
sessionInfo()
```

